\documentclass[11pt]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{array}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}

\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}
\newtheorem{conjecture}[definition]{Conjecture}
\newtheorem{remark}[definition]{Remark}

\newcommand{\TA}{\textnormal{\textsc{Traversal Algebra}}}
\newcommand{\trav}{\mathsf{trav}}
\newcommand{\emit}{\mathsf{emit}}
\newcommand{\reduceSrc}{\mathsf{reduceSrc}}
\newcommand{\reduceDst}{\mathsf{reduceDst}}
\newcommand{\denote}[1]{\lbrack\!\lbrack #1 \rbrack\!\rbrack}
\title{Traversal Algebra:\\A Pull--Map--Push Calculus for Parallel Graph Programs}
\author{Akira Hayakawa}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We develop \TA, a small algebra for expressing frontier-oriented graph
programs independently of a particular GPU execution strategy.  A traversal
pulls vertex and edge data to an ordered stream of edge contexts, a pure map
constructs messages, and monoidal terminals either emit the stream or push
messages back to source or destination vertices.  We fix the finite
ordered multigraph semantics and prove the foundational sequence and fusion
laws.  The definitions and theorems in this slice are checked in Lean~4, and
the executable Lean semantics generates conformance cases for the Massively
GPU implementation.  We additionally define Monoidal Frontier BSP as an
independent source machine and machine-check a forward simulation into the
pull--map--push semantic core for one superstep and every finite execution.
We then give a signature-relative, callback-free typed syntax and prove that
syntax-directed compilation commutes with denotation and preserves every
finite execution.  For the closed frontier-to-frontier fragment we define an
independently compositional \TA{} syntax with structural pulls and arbitrarily
nested map/zip trees.  We prove that sharing-aware normalization of every such
tree into a typed single-push normal form preserves pointwise values,
destination inboxes, every finite execution, and empty-frontier termination.
Combining normalization with translations to and from typed Monoidal Frontier
BSP establishes signature-relative semantic equivalence in both directions
for the complete closed expression language.  Explicit let bindings avoid the
exponential duplication possible with tree substitution.  A machine-checked
language-level cost semantics proves exact preservation of syntax size, unit
work, dependence depth, and local scalar storage, while proving non-increase
of full-stream temporary and materialization measures.  We also prove exact
order and multiplicity semantics for sparse candidate frontiers and give
explicit, bidirectionally equivalent observations for emission, source
reduction, and destination reduction.  A separate typed abstract CubeCL target
then derives execution and symbolic workgroup, subgroup, traffic,
synchronization, allocation, and materialization costs from the same closed
instructions, with exact linear or quasilinear terminal formulas.  The
Massively/CubeCL implementation is validated separately as an executable
artifact and is not an assumption of these language or abstract-machine
theorems.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-model}
\input{sections/03-calculus}
\input{sections/04-laws}
\input{sections/05-artifact}
\input{sections/06-completeness}

\bibliographystyle{plain}
\bibliography{references}

\end{document}
